---
title: "Introduction to R"
format: html
editor_options: 
  chunk_output_type: console
---

## Install packages

```{r}
#code chunk#1 
install.packages("tidyverse") #tidy data
install.packages("gtsummary") #create stats summary table
install.packages("labelled") #label variables
install.packages("skimr") #dataset summary
install.packages("haven") #import SPSS, Stata, SAS data
install.packages("readr") #import excel data
install.packages("qualtRics") #import Qualtrics data .csv
install.packages("Rcpp") #help to import data
```

## Load packages

tell R to load packages every session

```{r}
#code chunk#2
       
library(tidyverse)
library(gtsummary)
library(labelled)
library(skimr)
library(haven)
library(readxl)
library(qualtRics)
```

## Main Packages

tidyverse --- Tidy data

ggplot2 --- Grammar of Graphics

gtsummary --- Grammar of Tables

## Import data

Click File, Import Dataset, From Excel ...\

```{r}
#code chunk#3
covid_testing <- read_excel("covid_testing.xlsx")
View(covid_testing)
```

## R programming

```{r}
#code chunk#4
x <- 1+1
x
```

-   Your turn: change 1+1 to 4+6, run\

## R programming

object \<- function(x)

function(x)

## The pipe \|\> (%\>%)

x \|\>

function1 \|\>

function2 \|\>

function3

```{r}
#code chunk#5
covid_testing |>
  print() |>
  dim() # dimensions(#observations, #variables) 
```

## Inspect data

```{r}
#code chunk#6
glimpse(covid_testing)
```

## Inspect data

```{r}
#code chunk#7
skim(covid_testing)
```

## Data Management

Merge two datasets: covid_testing and covid_result

first import covid_result

```{r}
#code chunk#8
covid_result <- read_excel("covid_result.xlsx")
View(covid_result)
dim(covid_result)
```

## Merge

```{r}
#code chunk#9
covid_total <- full_join(covid_testing, covid_result, by="mrn")
view(covid_total)
dim(covid_total)
```

## Subset - select variables

Function: select

output \<- select(x, var1, var2, var3)

```{r}
#code chunk#10
data1<- select(covid_total, mrn, age, result)
print(data1)
dim(data1)
```

## Subset - exclude variables

output \<- select(x, -var1, -var2, -var3)

```{r}
#code chunk#11
data2<- select(covid_total, -mrn)
print(data2)
dim(data2)
```

## Subset - select observations

Function: filter

output \<- filter(x, argument)

```{r}
#code chunk#12
valid <- filter(covid_total, result != "invalid")
print(valid)
dim(valid)
```

## label variables

```{r}
#code chunk#13
covid_total <- 
  covid_total |>
   set_variable_labels(
     pan_day = "Day of the pandemic",
     age = "Age")
var_label(covid_total) #check variable labels
```

## Data type

-   continuous
-   categorical

## Visualize data - Histogram

```{r}
#code chunk#14
covid_total |>
  ggplot(aes(x= pan_day)) +
  geom_histogram()
```

## Visualize data - Histogram

```{r}
#code chunk#15
covid_total |>
  ggplot(aes(x= pan_day)) +
  geom_histogram(binwidth=1)
```

```{r}
#code chunk#16
covid_total |>
  ggplot(aes(x= pan_day, fill = result)) +
  geom_histogram(binwidth=1)
```

## Visualize data - Bar graph

For bar graphs, the x-axis variable is categorical

```{r}
#code chunk#17
covid_total |>
   ggplot(aes(x = result)) +
   geom_bar()
```

## Visualize data - Bar graph

```{r}
#code chunk#18
covid_total |>
   ggplot(aes(x = gender, fill = result)) +
   geom_bar()
```

## Visualize data - Scatter plot

Scatter plot of age vs time colored by gender, faceted by covid testing result

```{r}
#Code chunk#19
covid_total |>
  filter(result != "invalid") |>
  ggplot(aes(x = pan_day, y = age, color = gender)) +
  geom_point() +
  facet_grid(~result)
```

## Summary statistics

```{r}
#code chunk#20
covid_total |>
  tbl_summary()
```

## Summary statistics

exclude mrn

exclude variables that need to be grouped\

```{r}
#code chunk#21
covid_total |>
  select(-mrn, -clinic_name) |>
  tbl_summary()
```

## Comparing group demographics

Describe demographics between covid positive group and covid negative group

```{r}
#code chunk#22
covid_total |>
  select(-mrn, -clinic_name) |>
  filter(result != "invalid") |>
  tbl_summary(
    by = result
  )
```

## Bivariate analysis

```{r}
#code chunk#23
covid_total |> 
  select(-mrn, -clinic_name) |>
  filter(result != "invalid") |>
  tbl_summary(
    by = result
  ) |>
  add_p ()
```

## Your turn

create a summary table of age and result by gender\
dataset: covid_total\
variables: mrn, age, gender, clinic_name, result\

```{r}
#code chunk#24
object |>
  function(-var, -var) |>
  function(
    by = var
 )
```

## Alternative: Stata&R

Tidy data in Stata\
Import

```{r}
#code chunk#25
employee <- read_dta("employee.dta")
View(employee)
```

## Alternative: Stata&R

Create tables in R

```{r}
#code chunk#26
employee |>
  select(-id) |>
  tbl_summary()
```

```{r}
#code chunk#27 show stata labels
employee |>
  select(-id) |>
  mutate_at("jobcat", as_factor) |>
  tbl_summary()
```

## Missing values

-   suppress printing count of unknown values

```{r}
#code chunk#28
employee |>
  select(-id) |>
  tbl_summary(
      missing ="no"
)  #n=474
```

-   remove all observations with missing values

```{r}
#code chunk#29
employee |>
  select(-id) |>
  drop_na() |>
  tbl_summary()  #n=473
```

## Acknowledgement

<https://stephan-kadauke.quarto.pub/intro-to-r-for-clinical-data-rmed2022>

<https://www.danieldsjoberg.com/clinical-reporting-gtsummary-rmed>/

## Thank you for taking the time to learn R

Install R and RStudio on your computer

Install R packages

Learning by Doing

## Install R and RStudio

### For Windows users

[Download R for Windows](https://cran.rstudio.com/bin/windows/base/R-4.5.2-win.exe)

[Download RStudio for Windows](https://download1.rstudio.org/electron/windows/RStudio-2025.09.2-418.exe)

### For macOS users

[Download R for macOS 11 and higher](https://cran.rstudio.com/bin/macosx/big-sur-arm64/base/R-4.5.2-arm64.pkg)

[Download RStudio for macOS](https://download1.rstudio.org/electron/macos/RStudio-2025.09.2-418.dmg)

## Invite AI to the table

<https://umgpt.umich.edu/>

```{r}
str(covid_total)
```

Questions to UMGPT:

**Q1:** **I am working on a new dataset in R. I have an R object with the following data structure:**

```         
> str(covid_total) 
tibble [1,904 × 6] (S3: tbl_df/tbl/data.frame)  
$ mrn        : num [1:1904] 5e+06 5e+06 5e+06 5e+06 5e+06 ...  
$ gender     : chr [1:1904] "female" "female" "female" "male" ...  
$ pan_day    : num [1:1904] 32 29 29 25 21 27 34 32 31 32 ...  
$ clinic_name: chr [1:1904] "emergency dept" "clinical lab" "clinical lab" "clinical lab" ...  
$ age        : num [1:1904] 18 60 4 26 25 26 29 2 28 2 ...  
$ result     : chr [1:1904] "negative" "negative" "invalid" "negative" ...
```

**Can you tell me about it? Only summarize the data structure.**

**Q2:** **Help me write a R function that will check a data frame for NA values (missing data).**

**Q3: I want to know what the relationship between gender and covid testing results looks like. Can you help write R codes to visualize the relationship?**

**Q4: I** **would like to know if these relationships are statistically different, can you help me write R codes?**

**Q5:** **Help me interpret the output:**

**Q6: I got an error message:....... Here is my R codes:............Can you help explain and fix my R codes?**

**Q7:** **Can you exclude those that test is invalid, and then create a new variable that is 1 when result is "positive", but 0 otherwise?**

**Q8: I want to know......, then I want to know......, then I want to know......** **Ensure you understand the task. Explain it back to me before proceeding and I will check your logic.**

Other common questions:

Q1: How do I tell RStudio where it can find my files?

Q2: How do I load an Excel file into RStudio?

Q3: How do I save my data in R's native format?

Q4: How do I export my data?

## Additional codes: data management

```{r}
#rename variables: "new name" = "old name" 
covid_total <-
  covid_total |>
  rename("clinic" = "clinic_name") |>
  print(n=5)
```

```{r}
#reorder variables
covid_total |>
  select(mrn, age, gender, clinic, pan_day, result)
```

```{r}
#sort data
covid_total |>
  arrange(clinic, pan_day, result) |>
  print()
```

```{r}
#check data: what clinics included in this dataset
covid_total |>
  select(clinic) |>
  unique() |>
  print()
```

```{r}
#check in the first 30 days of pandemic, # of positive covid cases in each clinic
covid_total |>
  select(clinic, pan_day, result) |>
  filter(result != "invalid", pan_day <= 30) |>
  tbl_summary(
    by = result)
```

```{r}
#convert continuous variable "age" to categorical variable "age_group"
covid_total |>
  select(var=-mrn, -clinic) |>
  mutate(age_group = ifelse(age <= 18, "Youth", "Adult"),
      age_group = factor(age_group))|>
  tbl_summary(
    by=age_group
  )
```

```{r}
#convert character to numeric
#just showing the codes (pan_day was numeric)
covid_total |>
  mutate_at("pan_day", as.numeric)
```

```{r}
#create a publication-ready table
#change values: f,m to Female, Male
employee |>
  select(-id)|>
  mutate(gender = factor(gender,labels=             c("Female", "Male")))|>
  mutate_at("jobcat", as_factor) |>
  tbl_summary(
    label= list(gender ~ "Gender", 
           salbegin ~ "Beginning Salary (dollar)", salary ~ "Current Salary (dollar)", prevexp ~ "Previous Experience (months)"))
```
