---
title: ""
author: ""
date: 
format: html
execute:
  echo: false
  warning: false
editor: visual
---

## Title:

## Team

## Background

-   

## Study population

age

```{r}
#| echo: false
library(DiagrammeR)
grViz("digraph flowchart {
      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle]        
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']

      # edge definitions with the node IDs
      tab1 -> tab2 -> tab3 -> tab4 -> tab5;
      }

      [1]: 'N='
      [2]: 'N='
      [3]: 'Exclude N='
      [4]: 'Participants had complete data  N='
      [5]: 'Study population N='
      ")
```

## Objective

1.  Characterize

2.  Identify factors associated with

3.  

## Primary outcome

## Secondary outcome

## Variables of interest

-Demographic and clinical characteristics: race, ethnicity, bmi, gravida, para, insurance, primary language, social driver for health, comorbidities

-characteristics:

-characteristics:

## Data management (Import, Cleaning, Variable Creation)

### Load Packages

```{r}
library(tidyverse) #for import, cleaning, wrangling
library(gtsummary) #for summarizing
library(labelled) #for variable labelling
library(rio) #import/export 
library(janitor) #for cleaning and tabulation helpers
library(visdat) #data visualization for missingness
```

### Import and Merge data

```{r}
#--- Import raw datasets from REDCap ---
data <- read_csv("rawdata/EndometrialCancerGen_DATA_2025-09-15_0729.csv")
ruca <- read_csv("rawdata/Endometrial Cancer Genetic Counseling_FULL DATA_2025-09-03.csv")

#--- Select and format MRN and RUCA data ---
ruca <- ruca |>
  select(mrn, `Primary RUCA`) |>
  mutate(mrn = as.character(mrn))

#--- Merge RUCA to main data on MRN ---
data <- left_join(data, ruca, by = "mrn")
```

### Initial Data Check and Cleaning

```{r}
#--- Basic structure and missingness checks ---
str(data)          # Data structure
head(data)         # First 6 rows
dim(data)          # Number of rows & columns
names(data)        # Variable names
visdat::vis_miss(data) # Visualize missing data

#replace empty string with NA, whole dataset, all columns
data[data == ''] <- NA 
#replace only on selected columns
data <- data |> mutate_at('selected var', ~na_if(., '')) 

#--- Custom NA summary function ---
check_na <- function(df) {
  total_na <- sum(is.na(df))
  na_per_column <- colSums(is.na(df))
  list(total_na = total_na, na_per_column = na_per_column)
}
check_na(data)

#--- Duplicates ---
data |> get_dupes()
data |> get_dupes(mrn)
```

### **Variable Transformation and Labeling**

```{r}
#--- Recode variables for analysis ---
data <- data |>
  mutate(
    ethnic = factor(
      ethnic,
      levels = c(1:2, 990, 998),
      labels = c(
        "Hispanic or Latino",
        "Not Hispanic or Latino",
        "Prefer Not to Answer",
        "Unknown/Not Reported"
      )
    ),
    race = case_when(
      race___white == 1 ~ "White",
      race___black == 1 ~ "Black or African American",
      race___asian == 1 ~ "Asian",
      race___amind == 1 ~ "American Indian or Alaska Native",
      race___nahaw == 1 ~ "Native Hawaiian or Other Pacific Islander",
      race___other == 1 ~ "Other",
      race___unk == 1 ~ "Unknown/Not Reported",
      race___pnta == 1 ~ "Prefer Not to Answer"
    ),
    insurance_type = factor(
      insurance_type,
      levels = c(1:3, 5),
      labels = c(
        "Government", "Private/Commercial", "Uninsured/Self-Pay", "Unknown/Not Reported"
      )
    ),
    primary = case_when(
      primary_language == 1 ~ "English",
      primary_language %in% 2:10 ~ "Other",
      primary_language == 11 ~ "Unknown/Not Reported"
    ),
    tumor_histology = factor(
      tumor_histology,
      levels = c(1,3,4,5),
      labels = c("Endometrioid", "Clear Cell", "Carcinosarcoma", "Other")
    ),
    grade = factor(
      grade,
      levels = 1:4,
      labels = c("1", "2", "3", "Other/Not Reported")
    ),
    stage = case_when(
      stage___1 == 1  ~ "IA",
      stage___2 == 1  ~ "IB",
      stage___3 == 1  ~ "II",
      stage___4 == 1 | stage___5 == 1 | stage___6 == 1 | stage___7 == 1 ~ "IIIA-C2",
      stage___8 == 1 | stage___9 == 1 ~ "IVA-IVB",
      stage___999 == 1 ~ "Other/Not Reported"
    )
  )
```

### **Derived Variables for Timeline and Demographics**

```{r}
#--- Calculate time from diagnosis to genetic referral ---
data <- data |>
  mutate(
    date_of_diagnosis1 = ymd(date_of_diagnosis),
    date_of_referral_to_gc1 = ymd(date_of_referral_to_gc),
    time = as.numeric(date_of_referral_to_gc1 - date_of_diagnosis1, units = "days"),

    # Age at diagnosis in years
    date_of_birth1 = ymd(date_of_birth),
    age = as.duration(date_of_birth1 %--% date_of_diagnosis1) / dyears(1)
  )
```

### **Generate Cumulative Sums**

```{r}
#--- Summing social drivers and comorbidities for each row ---
data <- data |>
  rowwise() |>
  mutate(
    number_drive = sum(c_across(social_drivers_of_health___1:social_drivers_of_health___17), na.rm = TRUE),
    number_comor = sum(c_across(comorbidities___1:comorbidities___16), na.rm = TRUE)
  ) |>
  ungroup()
```

### **Numeric Conversion for Summaries**

```{r}
#--- Convert cumulative indexes to numeric ---
data <- data |>
  mutate(
    drive = as.integer(number_drive),
    comor = as.integer(number_comor),
    ruca = as.integer(`Primary RUCA`)
  )
```

### **Variable Grouping for Statistical Analysis**

```{r}
#--- Re-level and create grouped categorical/factor variables for analysis ---
data <- data |>
  mutate(
    race = as.character(race),
    ethnic = as.character(ethnic),
    race1 = if_else(race %in% c("Prefer Not to Answer", "Unknown/Not Reported"), "Not reported", race),
    ethnic1 = if_else(ethnic %in% c("Prefer Not to Answer", "Unknown/Not Reported"), "Not reported", ethnic),
    ruca1 = if_else(ruca %in% 1:3, "Metropolitan area", "Non-Metropolitan area"),
    drive1 = case_when(
      drive == 0 ~ "None",
      drive == 1 ~ "One",
      drive == 2 ~ "Two",
      drive >= 3 ~ ">=Three"
    ),
    comor1 = case_when(
      comor == 0 ~ "None",
      comor == 1 ~ "One",
      comor == 2 ~ "Two",
      comor >= 3 ~ ">=Three"
    ),
    histo = if_else(
      tumor_histology %in% c("Clear Cell", "Carcinosarcoma", "Other"), "Other", as.character(tumor_histology)
    ),
    stage1 = case_when(
      is.na(stage) | stage == "Other/Not Reported" ~ "Unstaged",
      TRUE ~ stage
    )
  )
```

### **Factor Level Orders & Additional Recoding**

```{r}
#--- Set factor reference/level orders for plotting/stats ---
data <- data |>
  mutate(
    race = forcats::fct_relevel(as.factor(race), "White", "Black or African American", "Asian", "Other", "Prefer Not to Answer", "Unknown/Not Reported"),
    ethnic = forcats::fct_relevel(as.factor(ethnic), "Not Hispanic or Latino", "Hispanic or Latino", "Prefer Not to Answer", "Unknown/Not Reported"),
    race1 = forcats::fct_relevel(race1, "White", "Black or African American", "Asian", "Other", "Not reported"),
    ethnic1 = forcats::fct_relevel(ethnic1, "Not Hispanic or Latino", "Hispanic or Latino", "Not reported"),
    drive1 = forcats::fct_relevel(drive1, "None", "One", "Two", ">=Three"),
    comor1 = forcats::fct_relevel(comor1, "None", "One", "Two", ">=Three")
  )

#--- Convert refer variable and set preferred level order ---
data <- data |>
  mutate(
    refer = factor(referral_to_gc, levels = 0:1, labels = c("No", "Yes")),
    refer = forcats::fct_relevel(refer, "Yes", "No"),
    result = forcats::fct_relevel(result, "Positive for mutation", "Negative for mutation", "Missing")
  )
```

### Additional Grouping

### **Variable Labelling**

```{r}
#--- Label variables for reporting/plotting ---
label(data$age)    <- "Age at diagnosis (yr)"
label(data$ethnic) <- "Ethnicity"
label(data$race)   <- "Race"
label(data$ethnic1)<- "Ethnicity"
label(data$race1)  <- "Race"
label(data$bmi)    <- "BMI"
label(data$primary)<- "Primary Language"
label(data$insurance_type) <- "Primary Insurance"
label(data$grade)  <- "Grade"
label(data$stage)  <- "Stage"
label(data$stage1) <- "Stage"
label(data$time)   <- "Time to Genetic Counseling Appointment (d)"
label(data$gravida) <- "Gravida"        
label(data$number_drive) <- "Social Drivers of Health" 
label(data$number_comor) <- "Comorbidities"
label(data$drive)  <- "Social Drivers of Health" 
label(data$comor)  <- "Comorbidities"
label(data$ruca)   <- "Primary RUCA"
label(data$drive1) <- "Social Drivers of Health" 
label(data$comor1) <- "Comorbidities"
label(data$ruca1)  <- "Primary RUCA"
label(data$tumor_histology) <- "Histology"
label(data$histo)   <- "Histology"
label(data$germline_testing_performed) <- "Prior Germline Testing"
label(data$completion_of_genetic_test) <- "Completion of genetic testing"
label(data$genetic_testing_result) <- "Genetic testing result" 
label(data$result) <- "Genetic testing result" 
label(data$refer)  <- "Referred to genetic counseling"
label(data$attended_gc_appt) <- "Attended Genetic Counseling Appointment"
label(data$location_of_gc_appt) <- "Location of Genetic Counseling Appointment"
label(data$recommended_cascade_test) <- "Recommendation for Cascade Testing"
label(data$concurrent_cancer) <- "Concurrent Cancer"
label(data$first_deg_relative_cancer) <- "First-Degree Relative with Cancer"
label(data$sec_deg_relative_cancer) <- "Second-Degree Relative with Cancer"
label(data$third_deg_relative_cancer) <- "Third-Degree Relative with Cancer"
label(data$gravida) <- "Gravida"
label(data$para) <- "Para"
label(data$para1) <- "Para"
```

### **Distribution & Missing Data Checks**

```{r}
#--- Normality tests ---
shapiro.test(data$age)     
shapiro.test(data$bmi)     
shapiro.test(data$comor)   
shapiro.test(data$drive)   
shapiro.test(data$para)    
shapiro.test(data$gravida) 

#--- List MRNs with missing info ---
data |>
  filter(is.na(third_deg_relative_cancer)) |>
  select(mrn) |>
  print(n = 100)
```

### **Save Final Data**

```{r}
#--- Save cleaned/curated dataframe ---
export(data, "data.rds")
```

## Descriptive summary

### Table 1 overview and by

```{r}
#| echo: false
library(tidyverse)
library(gtsummary)

# read rds
data <- readRDS("data.rds")
# descriptive summary; Wilcoxon rank-sum test, Chi-square test, Fisher's Exact test
data |>  
  select(age, bmi, race1, ethnic1, primary, ruca1,insurance_type, gravida, para1,drive1, comor1, histo, grade, stage1, refer, concurrent_cancer, germline_testing_performed) |>
  tbl_summary(
    by=refer,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_overall()|>
  add_p() |>
  modify_spanning_header(all_stat_cols() ~ "**Referred to Genetic Counseling**")|>
  modify_caption("**Data summary**")

#Manually calculate median, Q1, Q3
median(data$para, na.rm = TRUE)
quantile(data$para, probs = c(0.25, 0.75), na.rm = TRUE)
# Calculate median, Q1, and Q3 by group
summary_stats <- data %>%
  group_by(refer) %>%
  summarise(
    median_score = median(para, na.rm = TRUE),
    Q1 = quantile(para, 0.25, na.rm = TRUE),
    Q3 = quantile(para, 0.75, na.rm = TRUE)
  )

print(summary_stats)
# Run the Wilcoxon rank sum test
wilcox_test <- wilcox.test(para ~ refer, data = data)
print(wilcox_test)

```

## Multivariable Logistic Regression

### Outcome:

-   Multivariable logistic regression for estimated odds ratios (ORs) and 95% confidence intervals (CIs) for likelihood of outcome. Covariates include race/ethnicity, BMI, comorbidity, insurance status, etc.

    ```{r}
    glm(referral_to_gc ~ age + race1 + ethnic1 + bmi + ruca1 + primary + drive1 + comor1 + insurance_type + stage1, data=data, family = binomial) |>
    tbl_regression(exponentiate = TRUE) |>
    modify_caption("**Multivariable logistic regression**")

    ```

## Conclusion

## References

## Abstract

## Manuscript Discussions
