---
title: "R Codebook"
author: "Yang Liu"
date: 06-01-2023
format:
  revealjs:
    slide-number: c/t
    width: 1600
    height: 900
    css: ["theme/theme.css"]
    theme: simple
    echo: true   #include codes and results
    incremental: true
    scrollable: true
editor_options: 
  chunk_output_type: console
---
## identify multiple entries
```{r}
#identify duplicate entries
data |>
  get_dupes(mrn)
#subset data
dupdata <- data |>
  get_dupes(mrn) |>
  select(mrn, dupe_count, record_id, etdate, started_ocps, ohs_date, rpoc)  # Select only the columns you are interested in
#keep single entry per mrn
#first arrange the data frame by MRN then by palliative care consult in descending order (1, 0). 
#The distinct function is then used to remove duplicate rows. By specifying MRN as the variable for distinct to act upon, any further duplicates of that variable will be removed.
#The .keep_all = TRUE statement is to keep all other variables.
#data1 will only contain unique MRN entries where in case of duplicate MRN's, only the entry with palliative care consult is kept.
data1 <- data |>
  arrange(mrn, desc(`palliative_care_consult`)) |>
  distinct(mrn, .keep_all = TRUE)
```
## mrn Multiple entries-group_by
```{r}
#new variable "death_new": will be 1 for all rows of a patient who died at some point and 0 otherwise.
data <- data |> 
  group_by(mrn) |>
  mutate(death_new = as.integer(any(death_count2 == 1)))|>
  ungroup()
```


## Janitor
```{r}
data |> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
 janitor::convert_to_date(var)

```

##Replace empty cell with NA 

```{r}
# Replace empty cells with NA in dataset with base R
data[] <- lapply(data, function(x) {
  if (is.numeric(x) || is.integer(x)) {
    x[x == ""] <- NA
  } else if (is.character(x)) {
    x[x == ""] <- NA_character_
  } else if (is.factor(x)) {
    levels(x)[levels(x) == ""] <- NA
  }
  return(x)
})
#replace only on selected columns
data <- data |> mutate_at('var', ~na_if(., '')) 

```

##Replace NA with empty string 

```{r}
# whole dataset, all columns
covid_total[is.na(covid_total)] <- "" 

```

## replace values with NA

```{r}
data|>
  mutate(var=replace(var, var<=4, NA))
```

## recode NA

```{r}
#factor variable
data|>
  mutate(var=as.numeric(var))|>
  mutate(var=case_when(is.na(var) ~ 0,
                       TRUE ~ var))
#or
data |>
  mutate(var_clean=na_if(var, -99))
```
## how many NA
```{r}
data$var |> is.na() |>sum()
data$var |> is.na() |>which()
data$var |> anyNA()
data|> drop_na() |> nrow()
```
## visualize missing data
```{r}
library(visdat)
visdat::vis_miss(data)
#if dataset is too big
data |>
  select(var, .....)|>
  visdat::vis_miss()
```

## visualize missing data
```{r}
library(naniar)
gg_miss_var(data, show_pct = TRUE)
```

## add id
```{r}
data|>
  mutate(id = row_number())
```

## variable names
```{r}
library(janitor)
data <- data |>
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols")) #remove empty excel rows and columns
 janotor::convert_to_date(var) #convert excel dates to y-m-d

```
## data type
```{r}
typeof(data$var)#data type
newvar<- as.numeric(oldvar)
newvar<- as.character(oldvar)
newvar<- as.integer(oldvar)
```

## parse number
```{r}
data <- data |>
  mutate(var_clean=parse_number(var)) #input 20mg;output 20
```

## wide to long

```{r}
# pivot longer to gather _1, _2, _3, _4, _5 into a single column
df_long <- 
  data1|>
  select(id, antibiotic_name_id_1, antibiotic_name_id_2, antibiotic_name_id_3,antibiotic_name_id_4, antibiotic_name_id_5)|>
  pivot_longer(cols = starts_with("antibiotic_name_id_"),
               names_to = "antibiotic_name_id_col", 
               values_to = "antibiotic_name_id_val") 
#select(-c_col) # drop the original column names

# group by id and concatenate values into a space-separated string
df_new <- df_long %>%
  group_by(id) %>%
  summarize(c = paste(antibiotic_name_id_val, collapse = " ")) %>%
  ungroup() # remove the grouping
#add new variable to dataset        
data1<-
  data1 |>
  full_join(df_new, by=join_by(id))
```

## long to wide
```{r}
cmp1 <- cmp %>%
  select(category, study_cid)|>
  # Create a row number for each id which will serve as a suffix for category columns
  group_by(study_cid) %>%
  mutate(row = row_number()) %>%
  # Spread the data to wide format
  pivot_wider(names_from = row, values_from = category, names_prefix = "category") %>%
  # Ungroup the data frame
  ungroup()
```



## base R

```{r}
head(covid_total) # fist 6 rows
dim() #number of rows and columns
summary(covid_total) #basic descriptive
str(covid_total) #data structure
names(covid_total) #all variable names
boxplot.stats(covid_total$pan_day) #find outliers
shapiro.test(covid_total$pan_day) #Shapiro-Wilk normality test
```

## Data management {visibility="hidden"}

```{r}
#rename variables
covid_total <-
  covid_total |>
  rename("clinic" = "clinic_name") |>  #"new name" = "old name"
  print(n=10)
```

```{r}
#reorder variables
covid_total |>
  select(mrn, age, gender, clinic, pan_day, result)
```

```{r}
#sort data
covid_total |>
  arrange(clinic, pan_day, result) |>
  print()
```

```{r}
#check data: what clinics included in this dataset
covid_total |>
  select(clinic) |>
  unique() |>
  print()
```

```{r}
#check in the first 30 days of pandemic, # of positive covid cases in each clinic
covid_total |>
  select(clinic, pan_day, result) |>
  filter(result != "invalid", pan_day <= 30) |>
  tbl_summary(
    by = result)
```

## select columns
starts_with(“preop”) - matches preop_bp, preop_hr, preop_psa

ends_with(“pain”) - matches abd_pain, extremity_pain, wound_pain

contains(“bp”) - matches bp_screening, end_bp, first_bp_am

matches(“week[0-9]”) - matches week4, week8, week0

everything() - all columns

last_col() - the last column

num_range(“wk”, 1:6) - matches wk1, wk3, wk6, but not wk7

where(is.numeric) - selects variables of numeric type

```{r}
# select columns that are class Numeric
covid_total %>% 
  select(where(is.numeric)) %>% 
  names()
# select columns containing certain characters
covid_total %>% 
  select(contains("date")) %>% 
  names()
# searched for multiple character matches
covid_total %>% 
  select(matches("clinic|hosp|date")) %>%   # note the OR symbol "|"
# remove columns
covid_total %>% 
  select(-c(var1, var4:var6)) %>% # remove var1 and all columns from var4 to var6
  select(1:5)|> #select columns 1-5
  
```
## filter

```{r}
data |>
  filter(recurrence == 1)|>
  filter(preop_psa > 20)|>
  filter(near(preop_psa, 10))|>
  filter(near(preop_psa, 17, tol = 1.5)) #near a value with a set tolerance
  filter(between(preop_psa, 10, 13))|>
  filter(preop_psa %in% c(10,13,17))|>
  filter(str_detect(diagnosis, "oma"))|>#matches “oma”
  filter(!str_detect(diagnosis, pattern =  "Hodgkin"))|>
  filter(str_detect(diagnosis, pattern = "lympho.+ic"))|> #.+ means any character from 1 to N times
  filter(str_detect(diagnosis, "myelo.*"))|>#.* means any character from 0 to N times
  filter(between(fake_date, ymd(“2020-03-01”), ymd(“2020-03-31”)))|>#between 2 dates
  filter(fake_agvhd_date > today() - months(24))|>#24 months prior to today
  filter(wday(fake_bmt_date) %in% c(1,7))|> #filter to weekdays 1 or 7 (weekend)
  filter(is.na(fake_date))|>#filtering for the missing data in a variable
  filter (!is.na(fake_dx_date))|> #filter for the non-missing data in a variable
```
## join
```{r}
left_join(x=dataset1, y=dataset2, by="var")
right_join(x=dataset1, y=dataset2, by="var")
full_join(x=dataset1, y=dataset2, by="var")
inner_join(x=dataset1, y=dataset2, by="var")
test <- anti_join(x=full, y=training, by="var")
```

## slice
```{r}
data |>
  slice(100:200)|>#slice by position
  slice_max(age, n=20)|>#slice by top values in variable
  slice_min(pan_day, prop = 0.01)|>#slice by bottom values in variable
  slice_tail(prop = 0.01)#slice by top or bottom position
  slice_sample(prop = 0.3)|>#randomly samples 30 % obs
  slice_sample(n = 50)|>#randomly samples 50 obs
```

## remove duplicates

```{r}
covid_total |>
  distinct()  #evaluate and remove duplicate rows across all columns
```

## recode values

data \|\> mutate(var=recode(var, "old value"="new value")) \## logic criteria

```{r}
# Convert temperatures above 40 to NA 
data <- data %>% 
  mutate(temp = replace(temp, temp > 40, NA))
```

## ifelse example

```{r}
#IFELSE -convert continuous variable "age" to categorical variable "age_group"
covid_total |>
  select(var=-mrn, -clinic) |>
  mutate(age_group = ifelse(age <= 18, "Youth", "Adult"),
      age_group = factor(age_group))|>
  tbl_summary(
    by=age_group
  )
```
##convert character to numeric
```{r}

#just showing the codes (pan_day was numeric)
covid_total |>
  mutate_at("pan_day", as.numeric)
```

##convert factor to binary
```{r}
# no missing data
data %>%
    mutate(new_variable = if_else(old_variable == "none", 0, 1))

# missing data, keep NA
uti<- uti |>
  mutate(access=case_when(is.na(umich19) ~ NA_real_, 
                          umich19=="Never" ~ 1,
                          TRUE ~ 0))
```




## set specified factor level as reference in regression

```{r}
#forcats
data1$volume <- forcats::fct_relevel(data1$volume, "low")

#tidyverse
data1<- data1|> mutate(volume=factor(volume))|> relevel(ref="low") #didn't check

#base R
data1$volume <- relevel(data1$volume, ref="low")

```

## set specified factor level order
```{r}
#forcats
data1$volume <- forcats::fct_relevel(data1$volume, "low", "medium", "high")


data$race <- forcats::fct_infreq(data$race)
```

## group_by examples

```{r}
covid_total |>
  group_by(clinic)|>
  summarise(mean = mean(pan_day),
            sd=sd(pan_day))|>
  print()
```

```{r}
covid_total |>
  group_by(clinic, result)|>
  get_summary_stats(pan_day, type = "mean_sd")|>
  print()
```

## case_when examples

```{r}
data |>
  mutate(menopause1=case_when(menopause ==2 ~1,
                              menopause == 3 ~2,
                              TRUE ~ menopause))|> #numeric
  mutate(numre=case_when(is.na(num_re) ~ NA_character_,
    num_re %in% c("2", "18+","3", "5", "7")~ "2 or more", 
    TRUE ~ as.character(num_re)))|> #character
  mutate(cmp_pe=case_when(category_1 == 7 | category_2 == 7 |  category_3 == 7|category_4 == 7  |category_5 == 7|category_6 == 7
|category_7 == 7~1, TRUE ~ 0))
  
```

## label categorial variables

```{r}
data |>
   mutate(ethnicity =
           factor(
             ethnicity,
             levels = 0:1,
             labels = c("Hispanic or Latinx", "Not Hispanic or Latinx")),
         e_insurance_type=
           factor(
             e_insurance_type,
             levels = c(1:5,998:999),
             labels = c( "Commerical Insurance (non-HMO)", "Health Maintenance Organization","Government Insurance", "Uninsured/Self Pay without Insurance","Self Pay with Insurance", "Other", "Medicaid Patient"))) |>
   labelled::set_variable_labels(
    patient_id ="Patient ID number",
    dob="Date of Birth",
    ethnicity="Ethnicity")
```

## calculate age and duration

```{r}
data |>
  mutate(DOB = lubridate::ymd(dob),
       dvin_date = lubridate::ymd(dvin_dx_date),
        age_dvin = as.duration(DOB %--% dvin_date) / dyears(1),
        age_start = (age_presentation - yrs_dz),
       lsdvin = (age_dvin - age_start))
```

## rowwise: calculate the number of complications per id

```{r}
test<- test |>
  rowwise() |>
  mutate(number_cmp=sum (c_across(c(umich4_1:umich4_7))))
```
## rowsum
```{r}
data$uterine <- rowSums(data[, c("dandc", "uterinesurg")], na.rm = TRUE) 
```


## proportions with confidence intervals

frequency and CI, one decimal place

```{r}
# data source: IOL
data|>
  filter(eiol_final == 1 & nulli == 1)|>
  select(ga_grp_2, maternal_composite_new , cd, operative, chorioamnionitis, pph_1000, laceration_3or4, dystocia, maternal_icu)|>
  tbl_summary(
    by=ga_grp_2,
    missing = "no",
    digits = list(all_categorical() ~ c(0, 1))
  ) |>
  add_ci(pattern = "{stat} ({ci})",
         style_fun = everything() ~ purrr::partial(style_percent, digits = 1)) |>
  add_p()
```

## import and export 
```{r}
library(rio)
#import excel data
data <- read_excel("file path")
#import csv data
data <- read_csv("file path")
#import SPSS or SAS data
library(haven)
data <- read_sav("file path")
data <- read_sas("file path")
data <- read_dta("file path")
### IMPORTANT
# Convert haven_labelled columns to factors
data_converted <- data %>%
  mutate(
    across(where(is.labelled), as_factor)
  )
#export clean subset
export(data, "data.rds")
export(data, "data.csv")  
#read rds
data <- readRDS("data.rds")
```

## Publication-ready table

```{r}
employee |>
  select(-id)|>
  tbl_summary(
    label= list(jobcat ~ "Job category", 
           salbegin ~ "Beginning Salary (dollar)", salary ~ "Current Salary (dollar)", prevexp ~ "Previous Experience (months)"),
    by=gender,
    missing = "no",
    statistic = all_continuous() ~ "{mean} ({sd})",
    digits = list(all_categorical() ~ c(0, 1)),
    missing_text = "Missing") |># how missing values should display)
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2))|>
  add_overall() |>
  add_n() |>
  add_ci() |> #confidence intervals
  modify_header(label ~ "**Variable**") |>
  modify_header(
    stat_1 ~ "**Female n=200**",
    stat_2 ~ "**Male n=200**"
  ) |> 
  modify_spanning_header(all_stat_cols() ~ "**Current Employees**") |>
  modify_footnote(
    all_stat_cols() ~ "Median (IQR) or Frequency (%) 95% CI"
  ) |>
  modify_caption("**Table 1. Employee Characteristics**") |>
  bold_labels()
```

##calculate BMI {visibility="hidden"}

``` {r}
data |>
  mutate(height_cm=case_when(height_unit==1 ~ height_no*2.54,
                          height_unit==2 ~ height_no))|>
  mutate(weight_kg=case_when(weight_unit==1 ~ weight_no/2.205,
                          weight_unit==2 ~ weight_no))|>
  mutate(height_m=height_cm/100)|>
  mutate(bmi=weight_kg/height_m ^ 2)|>
  mutate(bmigroup = case_when(bmi < 18.5 ~ "Under weight",
                              bmi < 25 ~ "Normal weight",
                              bmi < 30 ~ "Overweight",
                              bmi >= 30 ~ "Obese"))|>
  mutate(bmi2 = ifelse( bmigroup=="Obese", ">=30", "<30"))
data$bmigroup<-  forcats::fct_relevel(data$bmigroup, "Under weight", "Normal weight", "Overweight","Obese")
  
```

##calculate surgeon volume by total cases {visibility="hidden"}

``` {{r}}
data source: MSQC
volume <-
  data|>
  select(physician_cid)|>
  group_by(physician_cid) %>%
  summarise(
    frequency = n()
  ) %>%
  ungroup() %>%
  summarise(
    median = median(frequency),
    q1 = quantile(frequency, 0.25),
    q2 = quantile(frequency, 0.5),
    q3 = quantile(frequency, 0.75),
    q4 = max(frequency)
  )
volume
volume |>
  group_by(physician_cid) %>%
  mutate(surgeon=case_when(frequency<4 ~1,
                           frequency<=12 ~2,
                           frequency<=33 ~3,
                           frequency>33 ~4))
```

##calculate surgeon volume by calendar year {visibility="hidden"}

``` {{r}}
data source: MSQC
s_all <-
  data|>
  select(physician_cid)|>
  group_by(physician_cid) |>
  summarise(
    frequency = n()
  )  
s2015 <-
  data|>
  filter(surgery_year==2015)|>
  select(physician_cid)|>
  group_by(physician_cid) |>
  summarise(
    frequency = n()
  ) |>
  ungroup() |>
  mutate(surgeon2015=case_when(frequency<6 ~"low",
                               frequency>=6 ~"high"))  
s2016 <-
  data|>
  filter(surgery_year==2016)|>
  select(physician_cid)|>
  group_by(physician_cid) |>
  summarise(
    frequency = n()
  ) |>
  ungroup() |>
  mutate(surgeon2016=case_when(frequency<12 ~"low",
                           frequency>=12 ~"high"))

s2017 <-
  data|>
  filter(surgery_year==2017)|>
  select(physician_cid)|>
  group_by(physician_cid) |>
  summarise(
    frequency = n()
  ) |>
  ungroup() |>
  mutate(surgeon2017=case_when(frequency<12 ~"low",
                               frequency>=12 ~"high"))

s2018 <-
  data|>
  filter(surgery_year==2018)|>
  select(physician_cid)|>
  group_by(physician_cid) |>
  summarise(
    frequency = n()
  ) |>
  ungroup() |>
  mutate(surgeon2018=case_when(frequency<12 ~"low",
                               frequency>=12 ~"high"))
test0 <-
  s2015 |>
  select(physician_cid, surgeon2015)|>
  full_join(s_all, s2015,by=join_by(physician_cid), relationship="many-to-many")

test <-
  s2016 |>
  select(physician_cid, surgeon2016)|>
  full_join(test0, s2016,by=join_by(physician_cid), relationship="many-to-many")

test1 <-
  s2017 |>
  select(physician_cid, surgeon2017)|>
  full_join(test, s2017,  by=join_by(physician_cid), relationship="many-to-many")

test2 <-
  s2018 |>
  select(physician_cid, surgeon2018)|>
  full_join(test1, s2018,  by=join_by(physician_cid), relationship="many-to-many")

data0 <-
  full_join(data, test2,  by=join_by(physician_cid), relationship="many-to-many")

data0<-
  data0 |>
  mutate(volume=case_when(c(surgery_year==2015 & surgeon2015=="low")~"low",
                          c(surgery_year==2015 & surgeon2015=="high")~"high",
                          c(surgery_year==2016 & surgeon2016=="low")~"low",
                          c(surgery_year==2016 & surgeon2016=="high")~"high",
                          c(surgery_year==2017 & surgeon2017=="low")~"low",
                          c(surgery_year==2017 & surgeon2017=="high")~"high",
                          c(surgery_year==2018 & surgeon2018=="low")~"low",
                          c(surgery_year==2018 & surgeon2018=="high")~"high"))
```

##convert duration datetime to time {visibility="hidden"}

``` {{r}}
#data source: MSQC, data format "1 days 03:36:00"
#surgery_time, operation time; 
#incision_time, time from incision to discharge
data |>
  select(surgery_duration, incision_to_discharge)|>
  mutate(surgeryt=str_sub(surgery_duration,8,15))|>
  mutate(surgery_duration1 = hms(surgeryt)) |> 
  mutate(surgery_duration2 = surgery_duration1 + days(str_sub(surgery_duration, 1, 1))) |>
  mutate(surgery_time = as.numeric(surgery_duration2, "hours"))|>  
  mutate(surgery_time = case_when(surgery_time < 0.5 ~ "NA",
                                  TRUE ~  as.character(surgery_time))) |>
  mutate(incision_to_discharge = case_when(
    incision_to_discharge %in% c("95 days 03:36:00", "94 days 00:48:00", "92 days 04:59:00") ~ "NA",TRUE ~ as.character(incision_to_discharge))) |>
  mutate(inci=str_sub(incision_to_discharge,8,15))|>
  mutate(inci1 = hms(inci)) |> 
  mutate(inci2 = inci1 + days(str_sub(incision_to_discharge, 1, 1))) |>
  mutate(incision_time = as.numeric(inci2, "hours")) |>
  mutate(incision_time = case_when(incision_time < 2.0 ~ "NA",
                                   TRUE ~  as.character(incision_time))) |>
  mutate_at(vars("surgery_time", "incision_time"), as.numeric)|>
  mutate(sameday=case_when(incision_time <=24 ~ 1,
                           incision_time >24 ~0,
                           TRUE ~ incision_time))|>
  mutate(sameday=factor(sameday))
```

## import qualtrics data

``` {{r}}
#import qualtrics data directly
#instructions:https://cran.r-project.org/web/packages/qualtRics/vignettes/qualtRics.html
install.packages("qualtRics")
library(qualtRics)
qualtrics_api_credentials(api_key = "n6Rh6tDRAvIqxn4pYmjzMV5MabN4hNU2ShMpWfa3", 
+                         base_url = "ca1.qualtrics.com",
+                         install = TRUE)
readRenviron("~/.Renviron")
surveys <- all_surveys()
mysurvey <- fetch_survey(
                    surveyID=surveys$id[4],                     verbose = TRUE)
view(mysurvey)
glimpse(mysurvey)
#never share API key
```

``` {{r}}
#import qualtrics data (.csv) to R
install.packages("qualtRics")
library(qualtRics)
mysurvey1<- read_survey("Q.csv")
view(mysurvey1)
```

```{r}
#import redcap (.csv) to R
```

##data visualization {visibility="hidden"}

```{r}
#plot the difference between two ggplot density distributions
library(ggplot2)
n_sp  <- 100000
n_dup <- 50000
D <- data.frame( 
     event=c(rep("sp", n_sp), rep("dup", n_dup) ), 
     q=c(rnorm(n_sp, mean=2.0), rnorm(n_dup, mean=2.1)) 
 )
ggplot( D, aes( x=q, y=..density.., col=event ) ) + geom_freqpoly()
D %>% group_by(event) %>% 
    # calculate densities for each group over same range; store in list column
    summarise(d = list(density(q, from = min(.$q), to = max(.$q)))) %>% 
    # make a new data.frame from two density objects
    do(data.frame(x = .$d[[1]]$x,    # grab one set of x values (which are the same)
                  y = .$d[[1]]$y - .$d[[2]]$y)) %>%    # and subtract the y values
    ggplot(aes(x, y)) +    # now plot
    geom_line()
#ref: https://stackoverflow.com/questions/42564106/how-to-plot-the-difference-between-two-ggplot-density-distributions?rq=1
```

## Homogeneity of Variance: F test
```{r}
f_test <- var.test(salary ~ male, data = data)
f_test
#interpretation: if p-value is less than 0.05, therefore we reject the null hypothesis and we give evidence our variance are heterogenous.
```

## Normality test
```{r}
shapiro.test(data$age) 
```

## Independent two samples t-test
```{r}
#Assumptions:
#The dependent variable is continuous  
#The observations are independent  
#The dependent variable is approximately normally distributed within each group  
#Homogeneity of variance  

# Normality by group
by(data$salary, data$gender, shapiro.test)

# Homogeneity of variance
#var.test() is the traditional F-test for comparing two variances, but it assumes the data are normally distributed. fligner.test() also tests whether variances are equal, but it is more robust when the data are skewed or contain outliers.
var.test(salary ~ gender, data = data)
fligner.test(salary ~ male, data = data)

# Independent two-samples t-test assuming equal variances
t.test(salary ~ gender, data = data, var.equal = TRUE)

# Welch's independent samples t-test (handle unequal variances)
t.test(salary_normal ~ male, data = data)
```

## Chi-squared test
```{r}
# Convert the two variables to factors if they are not already
data$var1 <- as.factor(data$var1)
data$var2 <- as.factor(data$var2)

# Run the chi-square test
chi_square_result <- chisq.test(data$var1, data$var2)

# View the result of the chi-square test
chi_square_result

```

## Fisher's Exact test
```{r}
# Create a contingency table
contingency_table <- table(data$var1, data$var2)
# Run Fisher's Exact Test
fisher_test_result <- fisher.test(contingency_table)
# if not enough work space: Fisher's Exact Test can be computationally intensive for large tables, and it requires a sufficiently large workspace to handle the calculation.
fisher_test_result <- fisher.test(contingency_table, workspace = 2e7, hybrid = TRUE)
# Print the results
print(fisher_test_result)
```
## Correlation
```{r}
library(corrr)
#--- Normality tests ---
shapiro.test(chyst$var_x)     
shapiro.test(chyst$var_y)  

#normal distribution
#Pearson's correlation (Pearson correlation (r), which measures a linear dependence between two variables (x and y). It’s also known as a parametric correlation test because it depends to the distribution of the data. It can be used only when x and y are from normal distribution. The plot of y = f(x) is named the linear regression curve.)
correlation_tab <- data %>% 
  select(var_x, var_y) %>%   # keep numeric variables of interest
  correlate()      # create correlation table (using default pearson)

correlation_tab    # print

#not normal distribution
#Spearman's rho (rank-based correlation coefficients,non-parametric)
correlation_tab <- data %>% 
  select(var_x, var_y) %>%   # keep numeric variables of interest
  correlate(method = "spearman")      # create correlation table (using spearman)

correlation_tab    # example: rho=0.535  moderately positive correlation

## remove duplicate entries (the table above is mirrored) 
correlation_tab <- correlation_tab %>% 
  shave()

## view correlation table 
correlation_tab

## plot correlations 
rplot(correlation_tab)
## base R
result <-cor.test(data$var_x, data$var_y,  method = "spearman")
result  #example: p <0.001 rho=0.535

#interpretation
#rho or r: Correlation coefficient is comprised between -1 and 1:
#-1 indicates a strong negative correlation : this means that every time x increases, y decreases 
#0 means that there is no association between the two variables (x and y) 
#1 indicates a strong positive correlation : this means that y increases with x 
#conclusion:
#We can conclude that var_x and var_y are significantly correlated with a correlation coefficient of 0.535 and p-value of less than 0.001.
```

## Calculate median, Q1 and Q3 
```{r}
median(subset$score_sc, na.rm = TRUE)
quantile(subset$score_sc, probs = c(0.25, 0.75), na.rm = TRUE)
```

## Wilcoxon Rank sum test
```{r}
# Calculate median, Q1, and Q3 by group
summary_stats <- subset %>%
  group_by(ace_new) %>%
  summarise(
    median_score = median(score_sc, na.rm = TRUE),
    Q1 = quantile(score_sc, 0.25, na.rm = TRUE),
    Q3 = quantile(score_sc, 0.75, na.rm = TRUE)
  )

print(summary_stats)
# Run the Wilcoxon rank sum test
wilcox_test <- wilcox.test(score_sc ~ ace_new, data = data)
print(wilcox_test)
```

## univariable logistic regression, unadjusted odds ratio

``` {{r}}
data1 |>
  select(var1, var2, var3, var4, var5) |>
  tbl_uvregression(method = glm, y=var5, exponentiate = TRUE)|>
  modify_caption("**Table. Univariable logistic regression**")
```

## Relative Risk (Incidence Rate Ratio)
```{r}
data1 |>
  select(var1, var2, var3, var4, var5) |>
  tbl_uvregression(method = glm, method.args = list(family = poisson),y=var5, exponentiate = TRUE, hide_n = TRUE)|>
  modify_caption("**Table. Poisson regression analysis of risk factors in relation to var5**")
```


## multivariable logistic regression

``` {{r}}
glm(var5 ~ var1 + var2 + var3 + var4, data=data1, family = binomial) |>
tbl_regression(exponentiate = TRUE) |>
modify_caption("**Multivariable logistic regression**")
```

## Multivariable linear regression
```{r}
#example:ResidentsResearch/Residents/2027/Gabrielle Schweitzen/msqc
m1 <- lm(log(val_surgtime) ~ uterine + val_age + race3 + ethnicity + insurance + bmi + sleep_apnea + hypertension + smoker +  copd + volume_new +  procedure + indication +priorpelvic_csection + priorpelvic_robotic +priorpelvic_open, data=data)
```


## mixed model
```{{r}}
glm(var5 ~ var1 + var2 + var3 +  (1 | var4), data=data1, 
    family=binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 10)|> 
tbl_regression(exponentiate = TRUE) |>
modify_caption("**Multivariable logistic regression**")
#var4: random effect variable, i.e. site, surgeon, hospital
```
## forest plot
```{r}
library(tidyverse)
library(forestplot)
#load data
nvss <- read_excel("C:/Users/ylyl/OneDrive - Michigan Medicine/ResidentsResearch/Fellows/Chris Hong/NVSS OASIS Tables 2023.06.13.xls", 
                   sheet = "aor")
#plot
nvss |>
  forestplot(labeltext = c(var, aor),
             clip = c(0.1, 15.2),
             boxsize = 0.2,
             xlog = TRUE) |>
  fp_set_style(box = "royalblue",
               line = "darkblue"
               ) |> 
  fp_add_header(var = c("", "Characteristics"),
                aor = c("", "aOR(95% CI)"))
  fp_set_zebra_style("#EFEFEF")
#ref: https://cran.r-project.org/web/packages/forestplot/vignettes/forestplot.html
```
## power/sample size-mean
```{r}
library(pwr)
results <- pwr::pwr.t.test(n = NULL, # note that n is per arm
                sig.level = 0.05, 
                type = "two.sample", 
                alternative = "two.sided", 
                power = 0.8, 
                d = 0.5 )
plot(results)
```
## power/sample size-two proportions
```{r}
library(pwr)
pwr.p.test(h = ES.h(p1 = 0.6, p2 = 0.66),
           n = NULL, # note that n is per arm
           sig.level = .05,
           power = 0.8,
           alternative = "two.sided")
```
## non-inferiority analysis-sample size
```{r}
# Parameters for sample size calculation
reference_treatment_efficacy <- 0.90
new_treatment_efficacy_min <- 0.75
non_inferiority_margin <- 0.15
alpha <- 0.05
power <- 0.80
delta <- reference_treatment_efficacy - new_treatment_efficacy_min - non_inferiority_margin

# Calculate the sample size
sample_size <- power.prop.test(p1 = reference_treatment_efficacy, 
                               p2 = reference_treatment_efficacy - non_inferiority_margin + delta, 
                               sig.level = alpha, 
                               power = power, 
                               alternative = "one.sided")

print(sample_size)
```
## non-inferiority test
```{r}
# Simulated data example based on calculated sample size
reference_n <- ceiling(sample_size$n)
new_n <- ceiling(sample_size$n)
reference_successes <- round(reference_n * reference_treatment_efficacy)
new_successes <- round(new_n * new_treatment_efficacy_min)

# Run non-inferiority test using prop.test (for comparing two proportions)
result <- prop.test(x = c(reference_successes, new_successes),
                    n = c(reference_n, new_n),
                    alternative = "greater",
                    correct = FALSE,
                    conf.level = 1-alpha)

print(result)
```



## model performance
```{r}
#linear regression model
library(flexplot)
model.comparison(model1, model2)
#logistic regression
library(easystats)
performance::check_model(model1)
performance::model_performance(model1)
```

## survey-weighted data
```{r}
library(haven)
library(tidyverse)
library(janitor)
library(gtsummary)
library(survey)
ir <- read_sav("C:/Users/ylyl/OneDrive - Michigan Medicine/ResidentsResearch/Fellows/Marie Bangura/FGC/SL_2019_DHS/SLIR7ASV/SLIR7AFL.SAV")
ir<- ir |>
  clean_names()
ir<-ir|>
  mutate(wt=v005/1000000,
# Complete case         
# NOTE: This creates a logical vector, taking on values TRUE and FALSE        
  nomiss = !is.na(v012) & !is.na(v101) & !is.na(v106) & !is.na(v130) & !is.na(v131) & !is.na(v025) & !is.na(v119) & !is.na(v121) & !is.na(v125) & !is.na(v136) & !is.na(v151) & !is.na(v171a) & !is.na(v190) & !is.na(v190a) & !is.na(g102))
# Complex sample design parameters
DHSdesign<-svydesign(id=~v021, strata=~v023, weights=~wt, data=ir)
#use the subset() function to create design objects that will produce analyses just for the cases with complete data. These design objects are not actually necessary for regression analyses, but will ensure that descriptive statistics are based on the same cases as regression analyses. 
DHSdesign.nomiss <- subset(DHSdesign, nomiss)
#demographic summary table
DHSdesign.nomiss |>
  tbl_svysummary(
    # stratify summary statistics by the "fgc" column
    by = g102,
    # summarize a subset of the columns
    include = c(v012, v101, v106,v130, v131,v025, v119, v121, v125, v136, v151, 
                v171a, v190, v190a, g102),
    # adding labels to table
    label = list(
      v012 ~ "Age",
      v106 ~ "Highest education level",
      v101 ~ "Region",
      v130 ~ "Religion",
      v131 ~ "Ethnicity",
      v025  ~   "Type of place of residence",
      v119  ~   "Household has: electricity",
      v121  ~   "Household has: television",
      v125  ~   "Household has: car/truck",
      v136  ~   "Number of household members (listed)",
      v151 ~    "Sex of household head",
      v171a ~   "Use of internet",
      v190  ~   "Wealth index combined",
      v190a  ~  "Wealth index for urban/rural"
    )
  ) |>
  add_p() |> # comparing values by "g102" column
  add_overall() |>
  # adding spanning header
  modify_spanning_header(c("stat_1", "stat_2") ~ "**FGC**")
#notes: Unweighted proportions reflect the composition of the sample, whereas weighted proportions estimate the population distribution.For categorical variables,  N and  n values in the table are estimated population totals – when we did unweighted analyses, these were sample sizes. ref: https://www.bookdown.org/rwnahhas/RMPH/survey-desc.html
```
## flow chart


## String detection-package "stringr"-- ifelse and case_when
```{r}
#code using package "stringr"
data <- data |>
  mutate(
    procedure = if_else(other_procedure_e == "{0}", 0, 1),
    indication = case_when(
      indication_for_procedure_e == "{6}" ~ "Only adeno",
      str_detect(indication_for_procedure_e, "\\b6\\b") & !str_detect(indication_for_procedure_e, "\\b7\\b") ~ "Adeno and other",
      indication_for_procedure_e == "{7}" ~ "Only aub/ fibroids",
      str_detect(indication_for_procedure_e, "\\b7\\b") & !str_detect(indication_for_procedure_e, "\\b6\\b") ~ "Aub/fibroids and other",
      str_detect(indication_for_procedure_e, "\\b6\\b") & str_detect(indication_for_procedure_e, "\\b7\\b") ~ "Adeno, aub, fibroid and other",
      TRUE ~ "Other"
    ),
    priorpelvicsurgery = if_else(prior_abdominopelvic_surgery_e == "{1}", 0, 1),
    priorpelvic_csection = if_else(str_detect(prior_abdominopelvic_surgery_e, "\\b2\\b"), "Yes", "No"),
    priorpelvic_robotic = if_else(str_detect(prior_abdominopelvic_surgery_e, "\\b3\\b"), "Yes", "No"),
    priorpelvic_open = if_else(str_detect(prior_abdominopelvic_surgery_e, "\\b4\\b|\\b5\\b"), "Yes", "No")
```

## missing data
```{r}
data<-data|>
  # Complete case         
# NOTE: This creates a logical vector, taking on values TRUE and FALSE        
  mutate(nomiss = !is.na(bmi) & !is.na(bmigroup))
```

## Functions
### check NA
```{r}
check_na <- function(df) {
  # Calculate total number of NA values in the data frame
  total_na <- sum(is.na(df))
  
  # Calculate the number of NA values for each column
  na_per_column <- colSums(is.na(df))
  
  # Create a summary list
  na_summary <- list(
    total_na = total_na,
    na_per_column = na_per_column
  )
  
  return(na_summary)
}
#example
check_na(mydata)
```

###remove categories that has 0% frequency
```{r}
# A function to drop levels with zero count in a factor variable
drop_zero_categories <- function(x) {
  # Convert to factor if not already
  x <- as.factor(x)
  # Keep only levels with non-zero count
  factor(x, levels = levels(x)[tabulate(x) > 0])
}

# Apply this function to specified columns
data <- data %>%
  mutate(
    race = drop_zero_categories(race),
    income = drop_zero_categories(income),
    education = drop_zero_categories(education),
    ethnicity = drop_zero_categories(ethnicity)
  )
```
## "any comorbidities"
```{r}
#create new variable "any_comorbidities"
data <- data %>%
  mutate(any_comorbidities = if_any(starts_with("comorbidities_category___"), ~ . == 1))   
```
## Table-remove all zero columns
```{r}
data |>
  select(where(~ any(. != 0))) |> # Exclude all-zero columns
   tbl_summary(
    digits = list(all_categorical() ~ c(0, 1)))
```

